---
title: "Which types of weather event are the most harmful to population health and the economy?"
author: "Phil Allen"
date: "25 January 2015"
output: html_document
---


## Synopsis

I interpreted "most damaging" to mean
"does the most total damage regardless of frequency". Other interpretations - 
such as damage relative to frequency of event, or events causing the most damage
in extreme cases - are possible, and produce different rankings.

From the specified data set, I chose a decade (1991 - 2000) in which
reporting of events seemed to be fairly established.

The initial exploration that led to these choices is partially documented 
in the data exploration section [below](#explore).

To get a single measure of impact on population health, I combined fatalities 
(FATALITIES column) and injuries (INJURIES column), weighted 100:1. I found the
event types with the top three impacts to be tornado, flood, excessive heat.

To get a measures of economic impact, I added the value of property damage 
(PROPDMG column) and crop damage (CROPDMG column). I found the top three event
types to be thunderstorm wind (code TSTM WIND), tornado, flash flood.


## Data Processing 

The National Oceanic and Atmospheric Administration (NOAA)
[describes](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
the data collected as follows:

a. The occurrence of storms and other significant weather phenomena having
sufficient intensity to cause loss of life, injuries, significant property damage,
and/or disruption to commerce;

b. Rare, unusual, weather phenomena that generate media attention, such as snow
flurries in South Florida or the San Diego coastal area; and

c. Other significant meteorological events, such as record maximum or minimum
temperatures or precipitation that occur in connection with another event.


```{r prereq, results='hide'}
#Pre-requisites
library("R.utils")
library("lubridate")
library("knitr")
library("ggplot2")
```
Load the data. Also, parse out the event dates, and turn their types into a factor,
to facilitate analysis.

```{r process, cache=TRUE}
dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
csvPath <- "data/StormData.csv"
bz2Path <- paste0(csvPath,".bz2")
if (!file.exists("data")) {
	dir.create("data")
}
if (!file.exists(bz2Path)) {
    download.file(dataURL, destfile = bz2Path)
}
if (!file.exists(csvPath)) {
	bunzip2(filename = bz2Path, destname = csvPath, remove = FALSE)
}
StormData <- read.csv("data/StormData.csv", stringsAsFactors = FALSE)
StormData$BGN_DATE <- strptime(StormData$BGN_DATE,"%m/%d/%Y", tz = "UTC")
StormData$END_DATE <- strptime(StormData$END_DATE,"%m/%d/%Y", tz = "UTC")
StormData$EVTYPE <- as.factor(StormData$EVTYPE)
```

Pick a decade in which reporting is well-established.

```{r choose, cache=TRUE}
decade <- StormData[year(StormData$BGN_DATE) > 1990
					& year(StormData$BGN_DATE) <= 2000,]
```

## Results

### Human harm

I decided to sum total harm by event type,
and to study both fatalities and injuries.

```{r human0, cache=TRUE}
decade.fat <- aggregate(FATALITIES ~ EVTYPE, data = decade, FUN = sum)
decade.inj <- aggregate(INJURIES ~ EVTYPE, data = decade, FUN = sum)
```


Intially consider fatalities. (I am not familiar with a
plot that conveys order as a well as a table.)

```{r fat, results='asis'}
kable(head(decade.fat[order(-decade.fat$FATALITIES),], n = 5),
	  row.names = FALSE,
	  col.names = c("Type of event", "Number of fatalities"),
	  caption = "Top five event types by number of fatalities")
```

The ordering for injuries is somewhat different.

```{r inj, results='asis'}
kable(head(decade.inj[order(-decade.inj$INJURIES),], n = 5),
	  row.names = FALSE,
	  col.names = c("Type of event", "Number of injuries"),
	  caption = "Top five event types by number of injuries")
```

So the rank of event types depends on the relative weight of these outcomes.
Many of the injuries will be life-changing,
but more will heal in a few weeks.
Arbitrarily choosing a ratio 100:1.

```{r human1, cache=TRUE}
decade.human.harm <- merge(decade.fat, decade.inj)
decade.human.harm$weighted =
	100 * decade.human.harm$FATALITIES + decade.human.harm$INJURIES
```

Thus, the weighted ranks are as follows.

```{r human2, results='asis'}
kable(head(decade.inj[order(-decade.inj$INJURIES),], n = 5),
	  row.names = FALSE,
	  col.names = c("Type of event", "Weighted human harm"),
	  caption = "Top five event types by weighted human harm")
```

### Economic impact

I decided to study the sum property and crop damage.

```{r damage0, cache=TRUE}
decade.damage <- aggregate(PROPDMG + CROPDMG ~ EVTYPE, data = decade, FUN = sum)
```

Here are the top ranked event types.

```{r damage, results='asis'}
kable(head(decade.damage[order(-decade.damage$PROPDMG),], n = 5),
	  row.names = FALSE,
	  col.names = c("Type of event", "Value of property and crop damage (K$)"),
	  caption = "Top five event types by value of damage")
```

## <a name="explore"/>Exploration of the data

### Distribution of impacts

There is evidence of very skew distribution in some key outcome variables,
as seen from the summaries and quantiles.

```{r skewness}
summary(StormData$FATALITIES)
quantile(StormData$FATALITIES, c(0.9, 0.99, 0.999, 0.9999))
summary(StormData$PROPDMG)
quantile(StormData$PROPDMG, c(0.9, 0.99, 0.999, 0.9999))
```

So comparing event types by their mean could be very misleading.

However, the impact variables are additive, so summing total impacts by event type
still makes sense.

### Changes in reporting over time

How has event reporting changed over the years?

```{r overtime}
StormData$count <- 1
counts <- aggregate(count ~ year(BGN_DATE), data = StormData, FUN = sum)
names(counts) <- c("Year", "Count")
qplot(Year, Count, data = counts, geom = "line") +
  scale_x_continuous(breaks = round(seq(min(counts$Year), max(counts$Year), by = 5),1)) +
  ggtitle("Growth in number of events reported")
```

The growth in the number of events seems to be dominated by changes in reporting,
rather than changes in weather.

It is likely that minor events were under-reported in the early years, so I chose
to miss early years out from analysis. Also, I wanted to choose a period before
climate change started to have measurable effects. In the end, I chose 1991 - 2000.

### Correlation of human and economic impacts?

Is there any strong correlation between fatalities and property damage, 
in the selected decade?
Take logs, because of the long tails seen above.

```{r corr}
p <- ggplot(data = decade, aes(log(1 + FATALITIES), log(1 + PROPDMG)))
p + geom_point() + ggtitle("Log plot of fatalities versus property damaged")
```

There's a range of property damage for no or low levels of fatalities.
Where there are fatalites, there is some correlation between fatalities 
and the value of property damage.
