---
title: "Which weather events are the most harmful?"
author: "Phil Allen"
date: "22 January 2015"
output: html_document
---


## Synopsis: 

Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

## Data Processing 

which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

```{r, cache=TRUE}
library("R.utils")
csvPath <- "data/StormData.csv"
bz2Path <- paste0(csvPath,".bz2")
if (!file.exists(bz2Path)) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = bz2Path)
}
if (!file.exists(csvPath)) {
  bunzip2(filename = bz2Path, destname = csvPath, remove = FALSE)
}
StormData <- read.csv("data/StormData.csv")
```



## Results



## Exploration of the source data

```{r}
names(StormData)
```

Linkd to doc explanation.

Outcome variables we care about are in columns: 

* People

** FATALITIES

** INJURIES

** PROPDMG

** PROPDMGEXP

** CROPDMG

** CROPDMGEXP

Factors to look for: 
* Meteorological measures - e.g. windforce, magnitude
* Geography - tornado alley?
* Time - are people better protected than they used to be?



Evidence of very skew distribution in some key variables.

```{r}
summary(StormData$FATALITIES)
quantile(StormData$FATALITIES, c(0.9, 0.99, 0.999, 0.9999))
summary(StormData$MAG)
quantile(StormData$MAG, c(0.9, 0.99, 0.999, 0.9999))
```


